Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-08-13 paper2.qmd
psychological methods, submission

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two key issues in the CJ literature. First, it critiques the reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce (BTL) model in the analysis of CJ data. Specifically, it argues that while the assumptions of equal discriminal dispersions and zero correlation simplify the trait measurement model, they may fail to capture the complexity of some traits or stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between trait measurement and statistical inference in CJ applications. It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.
  To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods. This extension integrates CJ's core theoretical principles alongside key assessment design features. It then translates these elements into a statistical model for analyzing dichotomous CJ data.
  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions. It also encourages researchers across the social sciences to adopt more robust and interpretable methodologies.
date: last-modified
bibliography: references.bib
---

@@

Drawing conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022]. The proposed approach is not an exception to this fundamental principle. Like all approaches grounded on causal inference, it relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025]. However, its purpose is not to deliver automatic answers when applied to a given CJ dataset. Instead, it encourages the formulation of precise questions and the explicit articulation of assumptions, fostering a generalizable understanding of the CJ system under study [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024]. This clarity is crucial because the accuracy of estimates and validity of inferences depend on how well the data and inferential goals align with a model's assumptions [@Kohler_et_al_2019]. Although this alignment remains to be empirically tested for the proposed models, the theory-driven nature of this approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], grounded in both established theory and existing evidence.

@@

**Materials and code availability:** The `CODE LINK` section at the top of the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/) provides access to all materials and code.

@@

**CRediT authorship contribution statement:** *Conceptualization:* J.M.R.E, T.vD., S.DM., and S.G.; *Methodology:* J.M.R.E, T.vD., and S.DM.; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* T.vD. and S.DM.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* T.vD., S.DM., and S.G.; *Visualization:* J.M.R.E.; *Supervision:* S.G. and S.DM.; *Project administration:* S.G. and S.DM.; *Funding acquisition:* S.G. and S.DM.

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: Daal
      non-dropping-particle: van
    orcid: 0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ) by addressing two key issues in the CJ literature. First, it critiques the reliance on Thurstone's Case V assumptions and the Bradley-Terry-Luce (BTL) model in the analysis of CJ data. Specifically, it argues that while the assumptions of equal discriminal dispersions and zero correlation simplify the trait measurement model, they may fail to capture the complexity of some traits or stimuli, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between trait measurement and statistical inference in CJ applications. It contends that while separating these processes simplifies the analysis, this practice can also undermine the resulting statistical inferences.
  To address these issues, this study extends Thurstone's general form using a systematic, integrated approach based on causal and Bayesian inference methods. This extension integrates CJ's core theoretical principles alongside key assessment design features. It then translates these elements into a statistical model for analyzing dichotomous CJ data.
  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, stressing the need for CJ models tailored to the experiments and data assumptions. It also encourages researchers across the social sciences to adopt more robust and interpretable methodologies.
date: last-modified
bibliography: references.bib
---

@@

Drawing conclusions from observed data always requires assumptions, whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022]. The proposed approach is not an exception to this fundamental principle. Like all approaches grounded on causal inference, it relies on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025]. However, its purpose is not to deliver automatic answers when applied to a given CJ dataset. Instead, it encourages the formulation of precise questions and the explicit articulation of assumptions, fostering a generalizable understanding of the CJ system under study [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024]. This clarity is crucial because the accuracy of estimates and validity of inferences depend on how well the data and inferential goals align with a model's assumptions [@Kohler_et_al_2019]. Although this alignment remains to be empirically tested for the proposed statistical model, the theory-driven nature of this approach provides a solid foundation for future empirical evaluation of its causal assumptions [@Deffner_et_al_2022], grounded in both established theory and existing evidence.

@@

**Materials and code availability:** A previous version of this manuscript, along with the associated materials and code (see the section titled `CODE LINK`), has been made publicly available at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).

@@

**CRediT authorship contribution statement:** *Conceptualization:* J.M.R.E, T.vD., S.DM., and S.G.; *Methodology:* J.M.R.E, T.vD., and S.DM.; *Software:* J.M.R.E.; *Validation:* J.M.R.E.; *Formal Analysis:* J.M.R.E.; *Investigation:* J.M.R.E; *Resources:* T.vD. and S.DM.; *Data curation:* J.M.R.E.; *Writing - original draft:* J.M.R.E.; *Writing - review and editing:* J.M.R.E., T.vD., S.DM., and S.G.; *Visualization:* J.M.R.E.; *Supervision:* S.G. and S.DM.; *Project administration:* S.G. and S.DM.; *Funding acquisition:* S.G. and S.DM.
